The Role of Ice Compositions and Morphology for Snowlines and the C/N/O Ratios in Active Disks

### 2016-02-21 tex/draft2/ms.tex
update

\begin{abstract}
The elemental compositions of planets define their chemistry, and could potentially be used as beacons for their formation location if the elemental gas and grain ratios of planet birth environments, i.e. protoplanetary disks, are well understood. In disks, the ratios of volatile elements, such as C/O and N/O, are regulated by the abundance of the main C, N, O carriers, the ice environment in which these carriers reside, and the presence of snowlines of major volatiles at different distances from the central star. We explore the effects of dynamical processes, molecular compositions and abundances, and the ice morphology of dust grains in disks on the snowline locations of the main C, O and N carriers, and their consequences for the C/N/O ratios in gas and dust throughout the disk. We find that radial drift and accretion alone can reduce the snowline radii of the main C, O and N carriers, i.e. H$_2$O, CO$_2$, CO and N$_2$, by 40-60\% compared to static disks. If CO and N$_2$ are bound to water ice instead of pure ices, their snowlines move inward by $\sim$$70$\%. Both of these effects substantially change the disk regions where C/O and N/O are enhanced over the stellar value. In the outer disk, the gaseous C/O and N/O are enhanced by factors of $\sim$$2$ and $\sim$$3$, respectively. Our estimates for the C/N/O ratios are only modestly affected by the presence of some C in the form of CH$_4$ and of some N in the form of NH$_3$.

@@

\begin {enumerate}
\item The chemical composition of protoplanetary disks is largely dictated by the freeze-out of volatile species, such as carbon, oxygen and nitrogen carriers.

\item Carbon and oxygen bearing molecules, such as H$_2$O, CO$_2$ and CO, as well as the C/O ratio in protoplanetary disks/giant planet atmospheres have been extensively studied from a theoretical standpoint (refs, Oberg+11, Ali-Dib+14, Madhusudhan+14, Molliere+15, etc.), and volatile snowlines have been detected in disks, such as H$_2$O and CO (refs). However, both observations and chemical models (refs) have shown that other volatiles are abundant in disks, for example nitrogen bearing species and hydrocarbons, thus shaping the disk composition.

\item Moreover, the snowline locations of these volatiles strongly depend on the grain morphology and the ice environment in which they reside. Laboratory experiments (cite Fayolle+16 and reference therein) have shown that volatiles such as CO and N$_2$ have significantly different binding energies depending on whether they are pure ices or layered on a water-ice substrate (reference Sections 3 and 4 for details).

\item Thus, there are other important snowlines in disks. Discuss Figure 1. Re-emphasize how this directly changes chemical abundances and C/N/O ratios throughout the disk, and thus the compositions of nascent giant planets.

\item We thus need a more thorough theoretical framework to understand how the processes above affect the C/O and N/O ratios in disks.  To address this issue, we expand the coupled drift-desorption model developed in \citet{piso15b} (hereafter Paper I) by considering additional volatile molecules and abundances, ice morphology, in addition to disk dynamical processes.

\end {enumerate}

@@

\section{CH$_4$ and C/O Ratios}
\label{sec:CH4}

@@

@@

Figure \ref{fig:CO_ratio} shows the H$_2$O, CO$_2$ and CO snowline locations for particles with initial sizes $\sim0.06$ cm $\lesssim s \lesssim$ 7 m as well as estimates for the C/O ratio in gas and dust in a viscous disk, with the CO snowline calculated under different grain morphologies as noted above. The true snowline for particles that desorb outside the static snowline is the static snowline itself, hence desorbing particles with $s<0.06$ cm do not form true snowlines. As calculated in Paper I, drift and gas accretion may move the snowlines inwards by up to 40-60\% compared to a static disk, and specifically by up to $\sim50$ \% in the case of the CO snowline. This result is preserved for the updated CO binding energies, both for pure ice and on a water ice substrate. However, a layered grain structure moves the CO snowline inwards significantly: for our fiducial disk model, $r_{\rm CO, layered} \approx 8.7$ AU. Thus if CO is bound to water ice instead of pure ices, the CO snowline may move inward by up to 70\%. This large inward movement of the CO snowline implies that C/O ratios of order unity may be reached much closer to the host star if CO is layered on a water ice substrate, and the CO snowline may be inside 10 AU for certain disk parameters.

@@

\section{Nitrogen and N/O Ratios}
\label{sec:N}

@@

As in the case for CO, the N$_2$ binding energy is also strongly dependent on the ice environment in which the grain resides. If N$_2$ is layered on a water ice substrate, the N$_2$ binding energy is 1266 K \citep{fayolle16}. Figure \ref{fig:NO_ratio} shows the H$_2$O, CO$_2$, CO and N$_2$ snowline locations in a viscous disk for particles with initial size $\sim0.06$ cm $\lesssim s \lesssim$ 7 m, and with the CO and N$_2$ snowlines calculated assuming different grain morphologies as explained above, as well as estimates for the N/O ratio throughout the disk. For simplicity, we assume that all nitrogen is bound in N$_2$. This choice is justified since the presence of some NH$_3$ only moderately changes the N/O ratio (see Figure \ref{fig:Nstatic}), and since we are primarily interested in the N$_2$ snowline locations rather than exact values for the N/O ratio. The innermost N$_2$ snowlines in the viscous disk, created by particles with $s \sim 7$ m for our fiducial model, are located at $r_{\rm N_2, pure} \approx 42$ AU for N$_2$ as pure ice and at $r_{\rm N_2, layered} \approx 11$ AU for N$_2$ layered on water ice. Compared to the static snowlines, i.e. $r_{\rm N_2, pure, static} \approx 79$ AU and $r_{\rm N_2, layered, static} \approx 25$ AU, drift and gas accretion move the snowlines inward by up to $\sim$50 \%, similar to the case for CO (see Section \ref{sec:CH4}). By comparing our results in the viscous disk, we find that the N$_2$ snowline moves inward by more than 70\% if N$_2$ resides in a water ice environment rather than as pure ice. Similarly to the CO case, the N$_2$ snowline may be close to or inward of 10 AU for certain disk models.

@@

>>>>

\begin{abstract}
The elemental compositions of planets define their chemistry, and could potentially be used as beacons for their formation location if the elemental gas and grain ratios of planet birth environments, i.e. protoplanetary disks, are well understood. In disks, the ratios of volatile elements, such as C/O and N/O, are regulated by the abundance of the main C, N, O carriers, the ice environment in which these carriers reside, and the presence of snowlines of major volatiles at different distances from the central star. We explore the effects of dynamical processes, molecular compositions and abundances, and the ice morphology of dust grains in disks on the snowline locations of the main C, O and N carriers, and their consequences for the C/N/O ratios in gas and dust throughout the disk. We find that radial drift and accretion alone can reduce the snowline radii of the main C, O and N carriers, i.e. H$_2$O, CO$_2$, CO and N$_2$, by 40-60\% compared to static disks. If CO and N$_2$ are bound to water ice instead of pure ices, their snowlines move inward by $\sim$$70$\%. Both of these effects substantially change the disk regions where C/O and N/O are enhanced over the stellar value, and taken together they may move the CO and N$_2$ snowlines by a factor of $\sim$7 inward. In the outer disk, the gaseous C/O and N/O are enhanced by factors of $\sim$$2$ and $\sim$$3$, respectively. Our estimates for the C/N/O ratios are only modestly affected by the presence of some C in the form of CH$_4$ and of some N in the form of NH$_3$.

@@

%\begin {enumerate}
The chemical composition of protoplanetary disks is largely dictated by the freeze-out of volatile species, such as carbon, oxygen and nitrogen carriers. The snowline locations of volatile molecules are crucial in determining disk chemical abundances in gas and dust, as well as planet compositions.

Carbon and oxygen bearing molecules, such as H$_2$O, CO$_2$ and CO, as well as the carbon-to-oxygen (C/O) ratio in protoplanetary disks and in giant planet atmospheres have been extensively studied from a theoretical standpoint (\citealt{oberg11}, \citealt{alidib14}, \citealt{madhu14}, \citealt{molliere15}), and the snowlines of volatiles such as H$_2$O and CO have been detected (\citealt{zhang13}, \citealt{qi13}). However, disk chemistry is complex, and it involves many species besides carbon and oxygen carriers \citep{henning13}. In particular, several other volatiles have been detected in disks, such as nitrogen bearing species and hydrocarbons (e.g., \citealt{mandell12}), which adds further complexity to the disk composition.

 %However, both observations and chemical models (refs) have shown that other volatiles are abundant in disks, for example nitrogen bearing species and hydrocarbons, thus shaping the disk composition.

Moreover, volatile snowline locations strongly depend on the grain morphology and the ice environment in which they reside. Laboratory experiments (\citealt{fayolle16} and reference therein) have shown that volatiles such as CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Sections \ref{sec:CH4} and \ref{sec:N}).

 It follows that there are several important snowlines in disks, determined both by the disk composition and ice morphology. Figure \ref{fig:CNOstatic} shows an example of the H$_2$O, CO$_2$, CO, CH$_4$, NH$_3$ and N$_2$ snowlines in a static disk, with CO and N$_2$ as pure and water dominated ices, in the top and bottom panel, respectively. The ordinate displays the total carbon, oxygen and nitrogen abundance in solids as function of the hydrogen abundance. We note that we display this Figure here for qualitative purposes; the quantitative choices that we make for volatile abundances are discussed in Sections \ref{sec:CH4} and \ref{sec:N}. As expected, the total grain abundance increases with semimajor axis, as more and more species sublimate. More importantly, the CO and N$_2$ snowlines move several tens of AU inward if the ices are water dominated rather than pure. This changes the chemical abundances both in gas and dust throughout the disk, directly affecting the compositions of nascent giant planets forming in situ.

% Disk dynamical processes such as radial drift of solids and viscous gas accretion onto the central store also change the position of the snowlines, as demonstrated by \citet{piso15b} (hereafter Paper I), which introduces another element of uncertainty in the snowline locations.

In order to explain this added chemical and physical complexity, a more thorough theoretical framework is needed. In this work, we expand the coupled drift-desorption model developed in \citet{piso15b} (hereafter Paper I) by considering additional volatile molecules and abundances, ice morphology, in addition to disk dynamical processes.

This paper is organized as follows. In Section \ref{sec:review}, we review the drift-desorption model developed in Paper I. We discuss the effect of different abundances of the main carbon and oxygen carriers, grain morphology and dynamical processes on the snowline locations of carbon and oxygen bearing species and the C/O ratio in Section \ref{sec:CH4}. In Section \ref{sec:N}, we perform the same analyses for the main nitrogen carriers and the nitrogen-to-oxygen (N/O) ratio. We summarize our findings in Section \ref{sec:summary}.

%\end {enumerate}

@@

\section{The Effect of CO Ice Morphology and CH$_4$ on the C/O Ratio}
\label{sec:CH4}

\subsection{Role of CH$_4$}

@@

\subsection{CO Ice Environment and Disk Dynamics}

@@

Figure \ref{fig:CO_ratio} shows the H$_2$O, CO$_2$ and CO snowline locations for particles with initial sizes $\sim0.06$ cm $\lesssim s \lesssim$ 7 m as well as estimates for the C/O ratio in gas and dust in a viscous disk, with the CO snowline calculated under different grain morphologies as noted above. The true snowline for particles that desorb outside the static snowline is the static snowline itself, hence desorbing particles with $s<0.06$ cm do not form true snowlines. As calculated in Paper I, drift and gas accretion may move the snowlines inwards by up to 40-60\% compared to a static disk, and specifically by up to $\sim50$ \% in the case of the CO snowline. This result is preserved for the updated CO binding energies, both for pure ice and on a water ice substrate. However, a layered grain structure moves the CO snowline inwards significantly: for our fiducial disk model, $r_{\rm CO, layered} \approx 8.7$ AU. Thus if CO is bound to water ice instead of pure ices, the CO snowline may move inward by up to 70\%. By taking into account both disk dynamics and of CO in a water dominated environment, we find that the CO snowline may move inward by a factor of $\sim$$7$ compared to a static disk with CO as pure ice.  This large inward movement of the CO snowline implies that C/O ratios of order unity may be reached much closer to the host star if CO is layered on a water ice substrate, and the CO snowline may be inside 10 AU for certain disk parameters.

@@

\section{The Effect of N$_2$ Ice Morphology and NH$_3$ on the N/O Ratio}
\label{sec:N}

\subsection{The Role of NH$_3$}

@@

\subsection{N$_2$ Ice Environment and Disk Dynamics}

As in the case for CO, the N$_2$ binding energy is also strongly dependent on the ice environment in which the grain resides. If N$_2$ is layered on a water ice substrate, the N$_2$ binding energy is 1266 K \citep{fayolle16}. Figure \ref{fig:NO_ratio} shows the H$_2$O, CO$_2$, CO and N$_2$ snowline locations in a viscous disk for particles with initial size $\sim0.06$ cm $\lesssim s \lesssim$ 7 m, and with the CO and N$_2$ snowlines calculated assuming different grain morphologies as explained above, as well as estimates for the N/O ratio throughout the disk. For simplicity, we assume that all nitrogen is bound in N$_2$. This choice is justified since the presence of some NH$_3$ only moderately changes the N/O ratio (see Figure \ref{fig:Nstatic}), and since we are primarily interested in the N$_2$ snowline locations rather than exact values for the N/O ratio. The innermost N$_2$ snowlines in the viscous disk, created by particles with $s \sim 7$ m for our fiducial model, are located at $r_{\rm N_2, pure} \approx 42$ AU for N$_2$ as pure ice and at $r_{\rm N_2, layered} \approx 11$ AU for N$_2$ layered on water ice. Compared to the static snowlines, i.e. $r_{\rm N_2, pure, static} \approx 79$ AU and $r_{\rm N_2, layered, static} \approx 25$ AU, drift and gas accretion move the snowlines inward by up to $\sim$50 \%, similar to the case for CO (see Section \ref{sec:CH4}). By comparing our results in the viscous disk, we find that the N$_2$ snowline moves inward by more than 70\% if N$_2$ resides in a water ice environment rather than as pure ice. Similarly to the CO case, the N$_2$ snowline may move by a factor of $\sim$$7$ inward if we account for both disk dynamics and ice morphology (water dominated compared to pure), and may be close to or inward of 10 AU for certain disk models.

@@

\item The locations of the CO and N$_2$ snowlines are uncertain when we take consider both viscous versus static disk, and pure versus water dominated ices. The snowlines in a viscous disk with CO in a water environment are by up to a factor of $\sim$7 closer to the host star that in a static disk with CO as pure ice.
